Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-06-02 paper/main.tex
paper

\abstract{
\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{andreas.weber@hhu.de}{andreas.weber@hhu.de}\\  
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.
}

@@

\section{Introduction}
The annotation of genes in DNA is fundamental for many downstream tasks, which 
depend on high quality annotations. Gene annotation can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, manual gene annotation is 
generally intractable. Instead, researchers can use pipelines such as 
Maker \citep{cantarel2008maker}, PASA \citep{haas2003improving} or those 
offered by genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl}. 
These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homologous sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State-of-the-art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble generalizing across species and it seems 
actual learning of underlying patterns is limited despite the fundamental ability 
to do so being present \citep{}.

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
Keras \citep{chollet2015keras} on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weigths as the class frequencies are both unbalanced and 
vary greatly between animals and plants. 
For example, the tradeoff between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

@@

The hyperparameters were optimized by a combination of manual and automatic 
optimization. Automatic optimization was carried out by using either the TPE 
algorithm \citep{bergstra2011algorithms}, random search or grid search depending on 
the situation. We used NNI \citep{nni2019} to facilitate the search. All relevant 
hyperparameters can be found in Table S\ref{suptab:params} and 
S\ref{suptab:classweights} or in the Helixer source code repository.

@@

\begin{table}[!t]
\processtable{Summary of Experimental Results\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 & 0.757 \\
Dilated CNN Baseline & 0.666 &  0.802 \\
Best Helixer model & 0.77  & 0.833  \\
Best model + varied input length & 0.834  & - \\
Best model (+ varied input length) + overlapping &  0.844  & 0.843  \\
(Varied input length +) overlapping + ensembling & 0.868  & 0.863  \\
\botrule
\end{tabular}}{{\it Note:} Values are the median in Subgenic F1 across all 
validation and test genomes of the respective group. Varied input length was only 
used in the animal case. The best model was chosen out of the eight models of the 
ensemble for having the best performance in the reported setup.}
\end{table}

@@

\begin{figure*}[!h]
\label{fig:cov_example_main}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_main}}
\caption{Four example helixer predictions in the context of RNAseq data, the reference,
and AUGUSTUS' prediction for {\it M. esculenta}. The examples were chosen so
that a) Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, b) Helixer had high accuracy but the 
reference was not supported by the RNAseq data, c) Helixer had low accuracy
and the reference was supported by the RNAseq data, and d) helixer had low
accuracy but the reference was not supported by RNAseq data. Plotability was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage 
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii) 
AUGUSTUS' predictions in matrix form, and iv) Helixer's predictions. The reference
and AUGUSTUS have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure*}

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, AUGUSTUS predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to AUGUSTUS. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer a), neither b), the reference, but not Helixer c),
and Helixer, but not the reference d).

@@

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c,S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (Figure S\ref{supfig:cov_example_02}d)
the RNAseq shows evidence of alternative splicing; 
and in another (Figure S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and of AUGUSTUS.

\begin{figure}[hbt]
\label{fig:average_vs_augustus}
\includegraphics[width=8.6cm]{images/cov_examples/average_vs_augustus}  % should be 8.6 when floats are sorted out right
\caption{Fraction of bp with color-indicated a) coverage and b) spliced coverage of genomic positions 
broken down by the confusion matrix of AUGUSTUS' and Helixer's predictions. Categories 
are only displayed if they can be meaningfully compared by examining a) coverage or b) spliced
coverage. The displayed fractions are the averages of the individual fractions for the 
six RNAseq-evaluation species. The left-most bars show cases where the two tools agree, 
while the remaining bars show paired conflicts. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.}
\end{figure}

@@

\section{Discussion}
With Helixer, we introduce a novel, deep-learning based tool for gene annotation
prediction. Helixer outperforms AUGUSTUS on base pair wise metrics and on 
consistency with independent RNASeq data.

We include trained models for plants and animals which achieve 
high prediction accuracy on cross-species gene annotation for broad 
phylogenetic groups (land plants and vertebrates, respectively).
Within these groups, this eliminates the dependency on retraining and the
expertise and data required therefore. Production of comparable, single-method
annotations for broad groups has the potential to greatly facilitate downstream
analyses; for instance it could avoid some of the inconsistency and errors that
are otherwise seen in RNAseq analyses when different annotations are used 
\citep{zhao2015comprehensive,torres2016robust}.

@@

\section*{Acknowledgements}

We are very grateful for the support of...

HHU HPC-team, deNBI, Ebenh{\"o}h lab for computational (particularly GPU) resources.

>>>>

\abstract{
\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/Helixer} under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{andreas.weber@hhu.de}{andreas.weber@hhu.de}\\  
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.
}

@@

\section{Introduction}
The annotation of genes in DNA is fundamental for many downstream tasks, which 
depend on high quality annotations. Gene annotation can be performed at different levels 
of precision, from simple coding -- non-coding classification to detailed structural 
labeling. Because of the sheer size of genomes alone, manual gene annotation is 
generally intractable. Instead, researchers can use pipelines such as 
Maker \citep{cantarel2008maker}, PASA \citep{haas2003improving} or those 
offered by genomic database providers like NCBI 
\citep{thibaud2013eukaryotic} and Ensembl \citep{aken2016ensembl}. 
These pipelines integrate experimental data (from e.g. RNA-seq 
or proteogenomics) with homologous sequences in the database and ab-initio gene 
predictions. The latter is an attractive approach, because it is cheap and fast. 
State-of-the-art performance is achieved by higher order hidden markov models 
(HMMs), such as Genscan \citep{burge1997prediction}, AUGUSTUS \citep{stanke2003gene} 
or SNAP \citep{johnson2008snap}. Their accuracy, however, leaves room for 
improvement. By encoding possible states and transitions in a probalistic model, 
designers of HMMs assume structure in the sequence that may limit its predictive 
power. In practice HMMs have trouble generalizing across species and any actual learning 
related to sequence motifs is limited to very short sequences that indicate state transitions.

@@

\subsection{Model Architecture and Training}
\label{sec:model}
We use a 4-layer deep stacked BLSTM network with 256 units per layer and layer 
normalization \citep{ba2016layer} between each BLSTM layer to produce the 
predictions in the form of a base pair wise classification. The model consists of 
circa 5.4 million parameters and was implemented with the deep learning library 
Keras (https://keras.io) on top of TensorFlow \citep{abadi2016tensorflow}. We 
tested multiple different architectures before arriving at this model configuration, 
including convolutional neural networks (CNN) and hybrid architectures. We also use 
class weigths as the class frequencies are both unbalanced and 
vary greatly between animals and plants. 
For example, the tradeoff between the number of intergenic and intronic bases is 
quite different in both groups (see Table \ref{tab:statistics_all}).

@@

The hyperparameters were optimized by a combination of manual and automatic 
optimization. Automatic optimization was carried out by using either the TPE 
algorithm \citep{bergstra2011algorithms}, random search or grid search depending on 
the situation. We used NNI (https://github.com/microsoft/nni) to facilitate the search. All relevant 
hyperparameters can be found in Table S\ref{suptab:params} and 
S\ref{suptab:classweights} or in the Helixer source code repository.

@@

\begin{table}[!t]
\processtable{Summary of Experimental Results\label{tab:results}} {
\begin{tabular}{@{}lll@{}}
\toprule & Animals & Plants\\
\midrule
AUGUSTUS & 0.632 & 0.757 \\
Dilated CNN Baseline & 0.666 &  0.802 \\
Best Helixer model & 0.77  & 0.833  \\
Best model + varied input length & 0.834  & - \\
Best model (+ varied input length) + overlapping &  0.844  & 0.843  \\
(Varied input length +) overlapping + ensembling & 0.868  & 0.863  \\
\botrule
\end{tabular}}{{\it Note:} Values are the median in Subgenic F1 across all 
validation and test genomes of the respective group. Varied input length was only 
used in the animal case. The best model was chosen out of the eight models of the 
ensemble for having the best performance on the validation set of the training genomes.}
\end{table}

@@

\begin{figure*}[!h]
\label{fig:cov_example_main}
\centerline{\includegraphics[width=\textwidth]{images/cov_examples/cov_example_main}}
\caption{Four example helixer predictions in the context of RNAseq data, the reference,
and AUGUSTUS' prediction for {\it M. esculenta}. The examples were chosen so
that {\bf a)} Helixer had high accuracy against the reference and the reference
was supported by the RNAseq data, {\bf b)} Helixer had high accuracy but the 
reference was not supported by the RNAseq data, {\bf c)} Helixer had low accuracy
and the reference was supported by the RNAseq data, and {\bf d)} helixer had low
accuracy but the reference was not supported by RNAseq data. Plotability was
a major secondary consideration. Each subplot shows from top to
bottom i) the natural log of the coverage (``cov", solid) and spliced coverage 
(``sc", dotted) + 1, ii) the reference annotation in matrix form, iii) 
AUGUSTUS' predictions in matrix form, and iv) Helixer's predictions. The reference
and AUGUSTUS have either 0 (white) or 1 (black) for each base pair and category, while
Helixer emits a probability from 0-1 represented via gray-scale. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.
}
\end{figure*}

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, AUGUSTUS predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to AUGUSTUS. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences (Figure \ref{fig:cov_example_main}) % and supplemental figures.
we identified cases where RNAseq supported a) both the annotations of the reference 
and Helixer, b) neither, c) the reference, but not Helixer,
and d) Helixer, but not the reference.

@@

Helixer models also sometimes showed uncertainty for larger regions.
In some cases, where Helixer did not receive RNAseq support for its highest probability
annotation, it assigned a low but non-trivial probability to the RNAseq-supported 
exon/intron pattern (Figures \ref{fig:cov_example_main}b-c, S\ref{supfig:cov_example_02}c). However, in the extreme,
there are cases where the Helixer model exibits substantial indecision or confusion 
and shifts gradually between classes with no single class receiving a high probability 
for extended stretches (Figures S\ref{supfig:cov_example_02}c-d). 
Notably, in one of the examples (Figure S\ref{supfig:cov_example_02}d)
the RNAseq shows evidence of alternative splicing; 
and in another (Figure S\ref{supfig:cov_example_02}c) Helixer's
prediction falls between that of the reference and of AUGUSTUS.

\begin{figure}[hbt]
\label{fig:average_vs_augustus}
\includegraphics[width=8.6cm]{images/cov_examples/average_vs_augustus}  % should be 8.6 when floats are sorted out right
\caption{Fraction of bp with color-indicated {\bf a)} coverage and {\bf b)} spliced coverage of genomic positions 
broken down by the confusion matrix of AUGUSTUS' and Helixer's predictions. Categories 
are only displayed if they can be meaningfully compared by examining a) coverage or b) spliced
coverage. The displayed fractions are the averages of the individual fractions for the 
six RNAseq-evaluation species. The left-most bars show cases where the two tools agree, 
while the remaining bars show paired conflicts. ``Ntrn" stand
for intron, and ``IG" stands for intergenic.}
\end{figure}

@@

\section{Discussion}
With Helixer, we introduce a novel, deep-learning based framework for the development 
of more capable tools for gene annotation prediction. 
Helixer outperforms AUGUSTUS on base pair wise metrics and on 
consistency with independent RNASeq data while also predicting cross-species for a wide range of  
genomes with one model.

We include trained models for plants and animals which achieve 
high prediction accuracy on gene annotation for broad 
phylogenetic groups (land plants and vertebrates, respectively).
Within these groups, this eliminates the dependency on retraining and the
expertise and data required therefore. Production of comparable, single-method
annotations for broad groups has the potential to greatly facilitate downstream
analyses; for instance it could avoid some of the inconsistency and errors that
are otherwise seen in RNAseq analyses when different annotations are used 
\citep{zhao2015comprehensive,torres2016robust}.

@@

\section*{Acknowledgements}
HHU HPC-team, deNBI, Ebenh{\"o}h lab for computational (particularly GPU) resources.
